---
title: "Extracting taxonomy information from metadata"
output:
  html_document:
    toc: false
    css: custom.css
---

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(ape)
library(metacoder)
library(knitr)
source("knitr_settings.R")
```

## Introduction

One of the most challenging parts of making inferences from publicly available DNA sequences for metabarcoding experiments is obtaining a set of reference sequences with reliable and specific taxonomy information. 
If the locus being studied is a standard barcoding locus and the taxa of interest well known, then there is likely a curated database to get sequences from. 
The problem becomes more difficult with non-standard loci or taxa. 
Some examples of curated databases are shown below:


Database                                                      | locus         | Taxon
------------------------------------------------------------- | ------------- | ------------
[UNITE](https://unite.ut.ee)                                  | ITS           | Fungi  
[RDP](http://rdp.cme.msu.edu/index.jsp)                       | 16S, 28S rRNA | Bacteria, Archea, Fungi
[ITS1](http://itsonedb.ba.itb.cnr.it:8080/ITS1)               | ITS1          | Fungi
[PR2](http://ssu-rrna.org/pr2)                                | 18S           | Protists
[SILVA](https://www.arb-silva.de/)                            | SSU, LSU      | Bacteria, Archaea and Eukarya
[greengenes](http://greengenes.lbl.gov/cgi-bin/nph-index.cgi) | 16            | Bacteria, Archaea


## Parsing taxonomy data with `extract_taxonomy`

There are a large number of sequence header/metadata formats and most of them are specific to a particular database.
This makes it difficult to compare and combine data from diverse sources. 
Rather than exacerbating the syntactic pollution with another custom format or creating a parser for every database-specific format, the function `extract_taxonomy` parses taxonomy information from arbitrary characters strings (e.g. sequence headers) identified by a regex expression.
Although the motivation for creating `extract_taxonomy` is to parse FASTA sequence headers, the function is not specific to FASTA or even to sequence information, so I will use the word "item" instead of "sequence header" from now on.
Any list of strings that contain identity and taxonomy information of a set of "items" is valid.

The function communicates with online databases (principally implemented using `taxize`) to infer missing information from information supplied.
For example, if a GenBank accession number was the only information available, the taxon id and classification would be retrieved from the NCBI databases.
At the minimum, the output of `extract_taxonomy` consists of unique taxon identifiers, and the tree structure of the taxonomy shared by all sequences.
This is all the information needed to fully characterize the taxonomic classification of a set of sequences.


### Input

There are three important arguments that will usually be relevant: `regex`, `key`, and `database`.

* The `regex` argument defines the structure of the strings and the location of relevant information (e.g. GenBank accession numbers) via regex capture groups (i.e. pairs of unescaped parentheses).
* The `key` argument defines what kind of information is in each capture group, determining how it will be parsed.
The elements of `key` has a defined set of possible values; see the `extract_taxonomy` help page for options. 
* The `database` argument determines the online taxonomy database that will be queried when necessary.
Usually, this is ncbi, but others databases are possible (although not as well tested yet); see the `extract_taxonomy` help page for options.

Other arguments are important in special cases.
Some will be explained in the examples below.

### Output

An object of type `taxmap` is returned.
See `?taxmap` for more details.

## Common usage examples

### Genbank FASTA headers

FASTA files downloaded from GenBank custom queries contain the GenBank id and the accession number/version to identify sequences.
Taxonomic information can be retrieved using either of these identifiers. 
The following shows how to extract the GI numbers, accession id, and description from the headers. 
The GenBank accession number is being used to look up the taxonomy information (hence `"item_id"` in `key` option), while the GI numbers and description are being returned without contributing taxonomic information (hence `"item_info"` in `key` option).

```{r eval=FALSE}
file_path <- system.file("extdata", "ncbi_basidiomycetes.fasta", package = "metacoder")
sequences <- ape::read.FASTA(file_path)
genbank_ex_data <- extract_taxonomy(sequences,
                                    regex = "^.*\\|(.*)\\|.*\\|(.*)\\|(.*)$",
                                    key = c(gi_no = "item_info", "item_id", desc = "item_info"),
                                    database = "ncbi")
```

The sequence headers have the format:

```{r, echo=FALSE, comment=NA}
head(item_data(genbank_ex_data, "input"))
```

We can plot the result using `plot`:

```{r}
plot(genbank_ex_data,
     node_size = n_items,
     node_color = n_items,
     node_label = name)
```


### UNITE FASTA headers

The format of the UNITE FASTA release has two pieces of information from which classifications can be determined.
The GenBank sequence identifier can be used to look up the taxon id from GenBank.
Alternatively, the classifications specified in the header can be used to make an arbitrarily coded taxonomy.

#### Using the sequence identifier

The GenBank accession number in the second entry of UNITE sequence headers can be used to look up the taxon assigned to each sequence by GenBank. 
Looking up the taxon assignment using the sequence accession number means changes in sequence taxonomy since the UNITE FASTA file was downloaded will be included. 
Therefore, the taxonomy returned by GenBank could be different than the one in the header.
However, some of the sequences in the UNITE database do not have a GenBank accession number; these IDs start with `UDB` and should be filtered out:

```{r eval=FALSE}
file_path <- system.file("extdata", "unite_general_release.fasta", package = "metacoder")
sequences <- ape::read.FASTA(file_path)
unite_ex_data_1 <- extract_taxonomy(sequences[!grepl(pattern = "\\|UDB", names(sequences))],
                            regex = "^(.*)\\|(.*)\\|(.*)\\|.*\\|(.*)$",
                            key = c(name = "item_info", "item_id",
                                    other_id = "item_info", tax_string = "item_info"),
                            database = "ncbi")
```

The sequence headers have the format:

```{r, echo=FALSE, comment=NA}
head(item_data(unite_ex_data_1, "input"))
```

```{r}
plot(unite_ex_data_1,
     node_size = n_items,
     node_color = n_items,
     node_label = name)
```

#### Using included classification names to look up the taxon id

If you want to use the structure and names of the classification provided in the header, but still look up the official taxon id, you can provide `"class"` as the only key with taxonomic information.
Since the UNITE classification also included the rank for each taxon, we will have to specify how to parse each taxon string using the `class_regex` and `class_key` options.
These work the same as `regex` and `key`, except they apply to each element in a classification.
You can still capture the sequence id or taxon id (assuming its present in the header) by using `"item_info"` or `"taxon_info"` where you would otherwise use `"item_id"` or `"taxon_id"`.

```{r eval=FALSE}
file_path <- system.file("extdata", "unite_general_release.fasta", package = "metacoder")
sequences <- ape::read.FASTA(file_path)
unite_ex_data_2 <- extract_taxonomy(sequences,
                                    regex = "^(.*)\\|(.*)\\|(.*)\\|.*\\|(.*)$",
                                    key = c(name = "item_info", sequence_id = "item_info",
                                            other_id = "item_info", "class"),
                                    class_regex = "^(.*)__(.*)$",
                                    class_key = c(unite_rank = "taxon_info", "name"),
                                    class_sep = ";",
                                    database = "ncbi")
```

```{r}
plot(unite_ex_data_2,
     node_size = n_items,
     node_color = n_items,
     node_label = name)
```

Note that the taxon name (entry 1) and the sequence id (entry 2) are now encoded as `"item_info"`, causing them to be interpreted as generic data.
This means that only the classification string (e.g. `k__Fungi;p__Ascomycota;c__Leotiomycetes`) will be interpreted as having taxonomic information, but the other information will also be included in the output in columns named `name` and `sequence_id`.
The unique taxon id for each taxon encountered will be looked up and taxa not found will be encoded as `NA`.

#### Using the included classifications to generate arbitrary ids

It is also possible to build a custom taxonomy encoding using the taxonomy in the sequence headers without looking up the official taxon ids of each taxon.
Taxa will be assigned arbitrary taxon ids that will be specific to the current analysis.
To do this, leave off the `database` option from the previous example.

```{r eval=FALSE}
file_path <- system.file("extdata", "unite_general_release.fasta", package = "metacoder")
sequences <- ape::read.FASTA(file_path)
unite_ex_data_3 <- extract_taxonomy(sequences,
                                    regex = "^(.*)\\|(.*)\\|(.*)\\|.*\\|(.*)$",
                                    key = c(seq_name = "item_info", sequence_id = "item_info",
                                            other_id = "item_info", "class"),
                                    class_regex = "^(.*)__(.*)$",
                                    class_key = c(unite_rank = "taxon_info", "name"),
                                    class_sep = ";")
```

```{r}
plot(unite_ex_data_3,
     node_size = n_items,
     node_color = n_items,
     node_label = name)
```

### ITS1 DB FASTA headers

The ITS1 database FASTA header includes the GenBank taxon id.
In the example below, both the `"item_id"` and `"taxon_id"` keys are provided, but only the `"taxon_id"` is used to look up taxonomy information since it has precedence. 

```{r eval=FALSE}
file_path <- system.file("extdata", "its1_chytridiomycota_hmm.fasta", package = "metacoder")
sequences <- ape::read.FASTA(file_path)
its1_ex_data <- extract_taxonomy(sequences,
                                 regex = "^(.*)\\|(.*)\\|tax_id:(.*)\\|(.*)$",
                                 key = c("item_id", taxon_name = "taxon_info",
                                         "taxon_id", description = "item_info"),
                                 database = "ncbi")
```

The sequence headers have the format:

```{r, echo=FALSE, comment=NA}
head(item_data(its1_ex_data, "input"))
```

```{r}
plot(its1_ex_data,
     node_size = n_items,
     node_color = n_items,
     node_label = name,
     layout = "fruchterman-reingold")
```

### PR2 FASTA headers

The first item in the header is sometimes a GenBank accession number, but not always.
Therefore, the best option is to use the included taxonomy information.
Note that there is no rank information available. 

```{r eval=FALSE}
file_path <- system.file("extdata", "pr2_stramenopiles_gb203.fasta", package = "metacoder")
sequences <- ape::read.FASTA(file_path)
pr2_ex_data <- extract_taxonomy(sequences,
                                regex = "^(.*\\..*?)\\|(.*)$",
                                key = c("item_id", "class"),
                                class_sep = "\\|")
```

The sequence headers have the format:

```{r, echo=FALSE, comment=NA}
head(item_data(pr2_ex_data, "input"))
```

```{r}
plot(pr2_ex_data,
     node_size = n_items,
     node_color = n_items,
     node_label = name)
```

### RDP FASTA headers

This RDP FASTA file does not contain any references to taxon or sequence ids that could be used to look up more information.
Instead, we will parse the taxonomy information included in the sequence headers.
In this case both the rank and taxon name are supplied.

```{r eval=FALSE}
file_path <- system.file("extdata", "rdp_current_Archaea_unaligned.fa", package = "metacoder")
sequences <- ape::read.FASTA(file_path)
rdp_ex_data <- extract_taxonomy(sequences,
                                regex = "^(.*?) (.*)\\tLineage=(.*)",
                                key = c(id = "item_info", description = "item_info", "class"),
                                class_regex = "(.+?);(.*?);",
                                class_key = c("name", "taxon_info"))
```

```{r, echo=FALSE, comment=NA}
head(item_data(rdp_ex_data, "input"))
```

Note that the same character ( `;` ) is used to delineate different ranks and taxa.
This makes it the `class_sep` option inappropriate.
If the `class_sep` option is not defined, each match of `class_regex` is considered to be a different taxa.
Therefore, we can use a `class_regex` that matches two `;` to correctly parse this awkward format.

```{r}
plot(rdp_ex_data,
     node_size = n_items,
     node_color = n_items,
     node_label = name,
     layout = "fruchterman-reingold")
```



### SILVA FASTA headers

```{r eval=FALSE}
file_path <- system.file("extdata", "silva_nr99.fasta", package = "metacoder")
sequences <- ape::read.FASTA(file_path)
silva_ex_data <- extract_taxonomy(sequences,
                                regex = "^(.*?) (.*)$",
                                key = c(id = "item_info", "class"),
                                class_sep = ";")
```


```{r}
taxon_data(silva_ex_data)
```


```{r}
plot(silva_ex_data,
     node_size = n_items,
     node_color = n_items,
     node_label = name,
     tree_label = name)
```



