---
title: "Manipulating taxonomic data"
output:
  html_document:
    toc: false
    css: custom.css
---

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(metacoder)
library(knitr)
source("knitr_settings.R")
```


## Introduction

Manipulating taxonomic data is a more subtle problem then it might seem at first.
For example, if you want to remove a taxon, do you remove its supertaxa and subtaxa as well?
What if there are sequences assigned to that taxon; are the sequences removed or reassigned to a preserved supertaxa?
What if the taxon is an internal node in the taxonomy; do you connect its supertaxa and subtaxa or break the taxonomy?
The answers to these questions depend on what the goal of the subsetting is.
MetacodeR uses `dplyr`-style functions for manipulating taxonomic data.
For each `dplyr` verb, there are two functions in metacodeR, one that manipulates the taxon portion of the data and one that manipulates the observation portion.

### Subsetting rows: `filter_taxa` and `filter_obs`

#### Removing supertaxa

Often, there are many shared ranks in the taxonomic hierarchy that can prevent effective visualization:

```{r}
plot(genbank_ex_data,
     node_size = n_obs,
     node_color = n_obs,
     node_label = name)
```

`filter_taxa` can easily remove these taxa by filtering by rank or selecting the root taxon by name:

```{r}
filter_taxa(genbank_ex_data, taxon_levels > 5) %>%
  plot(node_size = n_obs,
       node_color = n_obs,
       node_label = name)
filter_taxa(genbank_ex_data, name == "Basidiomycota", subtaxa = TRUE) %>%
  plot(node_size = n_obs,
       node_color = n_obs,
       node_label = name)
```


#### Removing subtaxa

You can also remove subtaxa by filtering with a specified level/rank:

```{r}
plot(unite_ex_data_3,
     node_size = n_obs,
     node_color = n_obs,
     node_label = name)
filter_taxa(unite_ex_data_3, taxon_levels <= 4) %>%
  plot(node_size = n_obs,
       node_color = n_obs,
       node_label = name)
```

Or we can filter by the number of observations assigned to each taxon:

```{r}
filter_taxa(unite_ex_data_3, n_obs >= 3) %>%
  plot(node_size = n_obs,
       node_color = n_obs,
       node_label = name,
       tree_label = name)
```

This is useful when plotting very large data sets, since it is difficult to make effective visualizations of over ~2000 taxa.
Note that observations assigned to removed subtaxa are reassigned to the closest supertaxa that passes the filter by default.
You can prevent this by setting the `reassign_obs` option to `FALSE`, but when most observations are assigned to tip taxa, this is rarely useful:

```{r}
filter_taxa(unite_ex_data_3, taxon_levels < 5, reassign_obs = FALSE) %>%
  plot(node_size = n_obs,
       node_color = n_obs,
       node_label = name)
```


#### Removing internal taxa

You can also remove internal taxa:

```{r}
filter_taxa(unite_ex_data_3, unite_rank != "c") %>%
  plot(node_size = n_obs,
       node_color = n_obs,
       node_label = name)
```

Note that the above result has no fungal classes anymore (taxa with names ending in "mycetes").
Like observations, subtaxa of removed taxa are reassigned to the closest supertaxon that passes the filter.
Although it usually does not make much sense to not reassign taxa, it is possible:

```{r}
filter_taxa(unite_ex_data_3, unite_rank != "c", reassign_taxa = FALSE) %>%
  plot(node_size = n_obs,
       node_color = n_obs,
       node_label = name)
```


#### Filtering observations 

Filtering observations assigned to taxa is less complicated.
The code below removes all taxa with `seq_id`s that do not start with "A".

```{r}
filter_obs(unite_ex_data_3, grepl("^A",  seq_id)) %>%
  plot(node_size = n_obs,
       node_color = n_obs,
       node_label = name)
```

You can remove any taxa that are made unobserved by the filtering using the `unobserved` option: 

```{r}
filter_obs(unite_ex_data_3, grepl("^A",  seq_id), unobserved = FALSE) %>%
  plot(node_size = n_obs,
       node_color = n_obs,
       node_label = name)
```


### Sampling rows: `sample_n_taxa` and `sample_n_obs`

Random sampling of taxa and observations is similar to subsetting, except you provide weights to each observation or taxon indicating how likely it is to be included in the subset.
The random subset of taxa or observations is then passed to `filter_taxa` or `filter_obs` respectively.
Therefore, all the options of `filter_taxa` or `filter_obs` can be used within `sample_n_taxa` and `sample_n_obs`.

#### Sampling observations

Sampling observations is useful for making a subset of a large data set (not that the example below uses a large data set):

```{r}
sample_n_obs(unite_ex_data_3, size = 100, unobserved = FALSE) %>%
  plot(node_size = n_obs,
       node_color = n_obs,
       node_label = name)
```

Weights can be assigned to observations to determine how likely each is to be sampled:

```{r}
sample_n_obs(unite_ex_data_3, size = 50, unobserved = FALSE,
               obs_weight = ifelse(grepl("Agaricales", seq_name), 100, 1)) %>%
  plot(node_size = n_obs,
       node_color = n_obs,
       node_label = name)
```

You can also assign weights to observations based on the taxon they are assigned to: 

```{r}
sample_n_obs(unite_ex_data_3, size = 100, unobserved = FALSE, 
               taxon_weight = 1 / n_obs) %>%
  plot(node_size = n_obs,
       node_color = n_obs,
       node_label = name)
```


#### Sampling taxa

Taxa can be sampled the same way observations are sampled.
The code below randomly selects 5 taxa of rank "class":

```{r}
set.seed(1)
sample_n_taxa(unite_ex_data_3, size = 5, subtaxa = TRUE, 
               taxon_weight = ifelse(unite_rank == "c", 1, 0)) %>%
  plot(node_size = n_obs,
       node_color = n_obs,
       node_label = name, tree_label = name)
```

When randomly subsetting taxa, pay special attention to the options of `filter_taxa` since excluding them can have a drastic effect:

```{r}
set.seed(1)
sample_n_taxa(unite_ex_data_3, size = 100) %>%
  plot(node_size = n_obs,
       node_color = n_obs,
       node_label = name)
```

#### Sampling a fraction of the total

`sample_n_taxa` and `sample_n_obs` have simple wrappers called `sample_frac_taxa` and `sample_frac_obs` that sample a given proportion of the total number of rows:

```{r}
set.seed(1)
sample_frac_obs(unite_ex_data_3, size = 0.1) %>%
  plot(node_size = n_obs,
       node_color = n_obs,
       node_label = name)
```


### Subsetting columns: `select_taxa` and `select_obs`

Subsetting columns is more straight forward that subsetting rows.
The functions `select_taxa` and `select_obs` are little more than wrappers for `dplyr::select`.
The only thing they do differently is to enforce that the `taxon_ids`, `supertaxon_ids`, and `obs_taxon_ids` columns are preserved:

```{r}
unite_ex_data_3
select_taxa(unite_ex_data_3, unite_rank)
```

Note how the "name" column has been removed from "taxon_data".
You can also use this to reorder columns: 

```{r}
select_obs(unite_ex_data_3, other_id, seq_name)
```


### Adding columns: `mutate_taxa` and `mutate_obs`

Adding a column to either `taxon_data` or `obs_data` is easy using the `dplyr` syntax:

```{r}
mutate_taxa(unite_ex_data_3, new_col = "Im new!")
```

A convenient attribute of `dplyr::mutate` is the ability to reference newly created columns:

```{r}
mutate_taxa(unite_ex_data_3,
            new_col = "Im new!",
            newer_col = gsub(pattern = "!", replacement = "er!!", new_col))
```

Adding observation columns with `mutate_obs` works the same way.


### Replacing columns: `transmute_taxa` and `transmute_obs`

`transmute_taxa` and `transmute_obs` make new columns and discard all of the old columns:

```{r}
transmute_taxa(unite_ex_data_3,
               new_col = "Im new!",
               newer_col = gsub(pattern = "!", replacement = "er!!", new_col))
```


### Reordering rows: `arrange_taxa` and `arrange_obs`

The way `taxmap` objects are defined, the order of all of the components do not matter.
This means its easy to reorder them to fit your needs:

```{r}
arrange_taxa(unite_ex_data_3, name)
```

To change the direction of ordering, use `dplyr::desc`:

```{r}
arrange_taxa(unite_ex_data_3, desc(name))
```
