---
title: Graphing the RDP database
---

```{r rdp_setup, echo=FALSE, warning=FALSE, message=FALSE}
source("settings.R")
```

## Requirements 

**NOTE:** This analysis requires at least 16Gb of RAM to run. 

## Parse database

The code below parses and subsets the entire RDP non-redundant reference database.
The object made is quite large.

```{r load_rdp}
file_path <- file.path(input_folder, "rdp_current_Bacteria_unaligned.fa")
system.time(rdp <- extract_taxonomy(seqinr::read.fasta(file_path, as.string = TRUE),
                                    regex = "\\tLineage(.*)",
                                    key = c("class"),
                                    class_regex = "[;=](.+?);(.+?)",
                                    class_key = c("name", rdp_rank = "taxon_info")))
print(rdp)
```

## Subset

```{r subset_rdp}
if (! is.null(max_taxonomy_depth)) {
  system.time(rdp <- filter_taxa(rdp, n_supertaxa <= max_taxonomy_depth))
  print(rdp)
}
if (! is.null(min_seq_count)) {
  system.time(rdp <- filter_taxa(rdp, n_obs >= min_seq_count))
  print(rdp)
}
if (just_bacteria) {
  system.time(rdp <- filter_taxa(rdp, name == "Bacteria", subtaxa = TRUE))
  print(rdp)
}
```



## Plot whole database

Although RDP is such a large database (`r nrow(rdp$taxa_data)` taxa) that graphing everything can be a bit overwhelming, it gives an intuitive feel for the complexity of the database:

```{r rdp_plot_all}
system.time(rdp_plot_all <- plot(rdp,
                             node_size = n_obs,
                             node_color = n_obs,
                             node_label = name,
                             node_label_max = 150,
                             node_color_axis_label = "Sequence count",
                             make_legend = FALSE,
                             output_file = file.path(output_folder, paste0("rdp--all", output_format))))
print(rdp_plot_all)
```


## PCR

```{r rdp_pcr}
system.time(rdp_pcr <- primersearch(rdp,
                                      forward = forward_primer,
                                      reverse = reverse_primer,
                                      mismatch = max_mismatch))
```

```{r rdp_plot_pcr_all}
system.time(rdp_plot_pcr_all <- plot(rdp_pcr,
                                       node_size = n_obs,
                                       node_label = name,
                                       node_color = prop_amplified,
                                       node_color_range =  pcr_success_color_scale,
                                       node_color_trans = "linear",
                                       node_label_max = 150,
                                       node_color_axis_label = "Proportion PCR success",
                                       node_size_axis_label = "Sequence count",
                                       output_file = file.path(output_folder, paste0("rdp--pcr_all", output_format))))
print(rdp_plot_pcr_all)
```

```{r, rdp_plot_pcr_fail}
system.time(rdp_plot_pcr_fail <- rdp_pcr %>%
              filter_taxa(prop_amplified < pcr_success_cutoff, supertaxa = TRUE) %>%
              plot(node_size = n_obs - count_amplified,
                   node_label = name,
                   node_color = prop_amplified,
                   node_color_range =  pcr_success_color_scale,
                   node_color_trans = "linear",
                   node_color_interval = c(0, 1),
                   node_label_size_range = c(0.008, 0.015),
                   node_label_max = 500,
                   node_color_axis_label = "Proportion PCR success",
                   node_size_axis_label = "Sequences not amplified",
                   make_legend = FALSE,
                   output_file = file.path(output_folder, paste0("rdp--pcr_fail", output_format))))
print(rdp_plot_pcr_fail)
```


## Save outputs for composite figure

Some results from this file will be combined with other to make a composite figure.
Below, the needed objects are saved so that they can be loaded by another Rmd file.

```{r rdp_save}
save(file = file.path(output_folder,"rdp_data.RData"),
     rdp_plot_all, rdp_plot_pcr_fail)
```

