---
title: "Untitled"
author: "Zachary Foster"
date: "June 24, 2016"
output: html_document
---

## flow chart code


```{r}
file_path <- system.file("extdata", "unite_general_release.fasta", package = "metacoder")
sequences <- ape::read.FASTA(file_path)
data <- extract_taxonomy(sequences,
                         regex = "^(.*)\\|(.*)\\|(.*)\\|.*\\|(.*)$",
                         key = c(seq_name = "obs_info", sequence_id = "obs_info",
                                 other_id = "obs_info", "class"),
                         class_regex = "^(.*)__(.*)$",
                         class_key = c(unite_rank = "taxon_info", "name"),
                         class_sep = ";")
```

```{r}
data %>%
  filter_taxa(name %in% c("Russula"), subtaxa = TRUE, supertaxa = TRUE, taxonless = FALSE, reassign_obs = FALSE) %>%
  filter_taxa(taxon_levels <= 6, taxonless = FALSE) %>%
  filter_taxa(! grepl("[_0-9]+", name), taxonless = FALSE, reassign_obs = FALSE) %>%
  arrange_taxa(c(p = 1, c = 2, o = 3, f = 4, g = 5, s = 6)[unite_rank])
plot(node_size = n_obs,
     node_color = n_obs,
     node_label = name,
     layout = "ka",
     make_legend = FALSE, 
     output_file = "figure_plot.pdf")
```



------------------ taxon_data -------------------
   taxon_ids parent_ids unite_rank           name
1          3         NA          p  Basidiomycota
2        305          3          c Agaricomycetes
3        323        305          o     Russulales
4        553        323          f    Russulaceae
5        557        553          g        Russula
6        575        557          s     chloroides
7        576        557          s    cyanoxantha
...      ...        ...        ...            ...

------------------ obs_data --------------------
   obs_taxon_ids            seq_name sequence_id
1             576 Russula_cyanoxantha    AB451977
2             578          Russula_sp    DQ424937
3             578          Russula_sp    HQ445312
4             578          Russula_sp    DQ061867
...      ...        ...        ...            ...



mutate_taxa, mutate_obs
transmute_taxa, transmute_obs
select_taxa, select_obs
filter_taxa, filter_obs
arrange_taxa, arrange_obs
sample_n_taxa, sample_n_obs




... f__Russulaceae;g__Russula;s__Russula_cyanoxantha
... f__Russulaceae;g__Russula;s__Russula_chloroides
... f__Russulaceae;g__Russula;s__Russula_delica
                       ...