
## Creating the HMP example dataset

### Sample data

```{r}
library(readr)
library(dplyr)
set.seed(1)
hmp_samples <- read_tsv("~/Downloads/v35_map_uniquebyPSN.txt")
hmp_otus <- read_tsv("~/Downloads/v35_psn_otu.genus.fixed.txt", skip = 1)
```

remove uneeded columns

```{r}
hmp_samples <- hmp_samples[, c("#SampleID", "sex", "HMPbodysubsite")]
```

rename columns

```{r}
colnames(hmp_samples) <- c("sample_id", "sex", "body_site")
```

subset treatments

```{r}
sites <- c("Saliva", "Throat", "Stool", "Right_Antecubital_fossa", "Anterior_nares")
hmp_samples <- hmp_samples[hmp_samples$body_site %in% sites, ]
```

rename treatments

```{r}
hmp_samples[hmp_samples$body_site == "Right_Antecubital_fossa", "body_site"] <- "Skin"
hmp_samples[hmp_samples$body_site == "Anterior_nares", "body_site"] <- "Nose"
```

remove samples not in abundance data

```{r}
hmp_samples <- hmp_samples[hmp_samples$sample_id %in% colnames(hmp_otus), ]
```

subsample sites

```{r}
hmp_samples <- hmp_samples %>%
  group_by(body_site, sex) %>%
  sample_n(size = 5)
```

convert sample IDs to character

```{r}
hmp_samples$sample_id <- as.character(hmp_samples$sample_id)
```


### Abundance matrix

Subset samples

```{r}
hmp_otus <- hmp_otus[, c("#OTU ID", "Consensus Lineage", hmp_samples$sample_id)]
```

rename columns

```{r}
colnames(hmp_otus)[1:2] <- c("otu_id", "lineage")
```


Remove OTUs with missing info

```{r}
hmp_otus <- hmp_otus[! endsWith(hmp_otus$lineage, "__"), ]
hmp_otus <- hmp_otus[! grepl(hmp_otus$lineage, pattern = "__;", fixed = TRUE), ]
```

Remove ambiguous OTUs

```{r}
hmp_otus <- hmp_otus[! grepl(hmp_otus$lineage, pattern = "IncertaeSedis"), ]
```

Remove singletons

```{r}
hmp_otus <- hmp_otus[rowSums(hmp_otus[, hmp_samples$sample_id]) > 1, ]

```


random subsample of OTUs

```{r}
hmp_otus <- hmp_otus %>% sample_n(1000, weight = rowSums(hmp_otus[, hmp_samples$sample_id]))
```

Add root rank 


```{r}
hmp_otus$lineage <- paste0("r__", hmp_otus$lineage)
```

## Add to package

```{r}
devtools::use_data(hmp_otus, overwrite = TRUE)
devtools::use_data(hmp_samples, overwrite = TRUE)
```



## How to parse

```{r}
x = parse_tax_data(hmp_otus, class_cols = "lineage", class_sep = ";",
               class_key = c(tax_rank = "info", tax_name = "taxon_name"),
               class_regex = "^(.+)__(.+)$")
```


```{r}
library(metacoder)
heat_tree(x, node_size = n_obs, node_color = n_obs, node_label = taxon_names)
```

